---
title: "Exploratory Data Analysis of the Instagram Reach Dataset"
subtitle: "EDA, feature engineering, and interpretable classification for high-engagement Instagram posts"
author: "Sotiris Kylintireas — 3220098"
date: "2026-03-17"
always_allow_html: true
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: cosmo
    df_print: paged
    highlight: tango
    fig_caption: yes
    fig_width: 8
    fig_height: 5
  word_document:
    toc: true
    toc_depth: '2'
  pdf_document:
    toc: true
    toc_depth: '2'
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  out.width = "90%"
)

suppressPackageStartupMessages({
  library(knitr)
  library(kableExtra)
  library(ggplot2)
  library(dplyr)
  library(scales)
  library(pROC)
})

set.seed(42)
options(scipen = 999)

theme_report <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, color = "grey40"),
      axis.title    = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}
```

```{r load-and-prepare, include=FALSE}
instagram_reach <- read.table(
  "data/instagram_reach.txt",
  header = TRUE,
  stringsAsFactors = FALSE
)

instagram_reach <- instagram_reach[, -1]

instagram_reach$USERNAME <- as.factor(instagram_reach$USERNAME)
instagram_reach$Caption  <- as.character(instagram_reach$Caption)
instagram_reach$Hashtags <- as.character(instagram_reach$Hashtags)

# Feature engineering: time category
instagram_reach$time <- ifelse(instagram_reach$Hours.since.posted == 2, 2,
                        ifelse(instagram_reach$Hours.since.posted == 3, 3, 4))
instagram_reach$time <- as.factor(instagram_reach$time)

# Feature engineering: like tier (quartile-based)
Q1 <- quantile(instagram_reach$Likes, 0.25)
Q3 <- quantile(instagram_reach$Likes, 0.75)
instagram_reach$like <- ifelse(instagram_reach$Likes < Q1, 1,
                        ifelse(instagram_reach$Likes <= Q3, 2, 3))
instagram_reach$like <- as.factor(instagram_reach$like)

# Binary target: high engagement
instagram_reach$high_like <- factor(
  ifelse(instagram_reach$like == 3, 1, 0),
  levels = c(0, 1),
  labels = c("Low/Medium Engagement", "High Engagement")
)

# Train/test split (70/30)
train_idx  <- sample(nrow(instagram_reach), size = floor(0.70 * nrow(instagram_reach)))
train_data <- instagram_reach[train_idx, ]
test_data  <- instagram_reach[-train_idx, ]

# Logistic regression trained on training set
model_logit <- glm(high_like ~ Followers + Hours.since.posted,
                   data   = train_data,
                   family = binomial)

# Predictions on held-out test set
predicted_prob  <- predict(model_logit, newdata = test_data, type = "response")
predicted_class <- ifelse(predicted_prob > 0.5, 1, 0)
actual_class    <- as.numeric(test_data$high_like) - 1

# Confusion matrix
cm <- table(Predicted = predicted_class, Actual = test_data$high_like)

# Safe extraction from confusion matrix
get_cm_val <- function(cm_table, pred_label, actual_label) {
  pred_str <- as.character(pred_label)
  if (pred_str %in% rownames(cm_table) && actual_label %in% colnames(cm_table)) {
    cm_table[pred_str, actual_label]
  } else {
    0
  }
}

tp <- get_cm_val(cm, 1, "High Engagement")
fp <- get_cm_val(cm, 1, "Low/Medium Engagement")
fn <- get_cm_val(cm, 0, "High Engagement")
tn <- get_cm_val(cm, 0, "Low/Medium Engagement")

accuracy  <- mean(predicted_class == actual_class)
precision <- if ((tp + fp) > 0) tp / (tp + fp) else NA
recall    <- if ((tp + fn) > 0) tp / (tp + fn) else NA
f1        <- if (!is.na(precision) && !is.na(recall) && (precision + recall) > 0)
               2 * precision * recall / (precision + recall) else NA

roc_obj <- roc(actual_class, predicted_prob, quiet = TRUE)
auc_val <- as.numeric(auc(roc_obj))

# Baseline: always predict majority class (Low/Medium Engagement)
baseline_accuracy <- mean(actual_class == 0)

# Full-dataset model (for stable coefficient display)
model_full <- glm(high_like ~ Followers + Hours.since.posted,
                  data   = instagram_reach,
                  family = binomial)
```

## Project Objective

This project analyzes a dataset of 100 Instagram posts to identify the variables most associated with high engagement. The analysis explores the role of follower count and posting age in determining whether a post receives an above-average number of likes, and evaluates this relationship through a logistic regression classifier.

**Variables explored:** `Followers`, `Hours.since.posted`, `Likes`, along with engineered categorical features `time`, `like`, and `high_like`.

**Modeling task:** Binary classification — predicting whether a post falls in the high-engagement category based on follower count and time since posting.

## Why This Project Matters

Engagement prediction is a practical problem for anyone managing social media content. Understanding which factors drive higher engagement allows content creators and analysts to make data-informed decisions about posting strategy. This project demonstrates that simple, interpretable models built on a minimal feature set can already reveal meaningful patterns — a useful baseline before investing in more complex approaches. It also illustrates why thorough exploratory analysis before modeling is important: distributional asymmetries, class imbalance, and variable relationships all affect how results should be interpreted.

## Executive Summary

**Dataset:** 100 Instagram posts, 6 variables after cleaning.
**Goal:** Identify engagement patterns and predict high-engagement posts.
**Key findings:** Posting age (`Hours.since.posted`) is a stronger predictor than follower count. Posts in the highest time category show notably higher engagement rates. The `Followers` distribution is strongly right-skewed.
**Model result:** Trained on 70 posts and evaluated on a held-out test set of 30 — accuracy `r round(accuracy * 100, 1)`%, AUC `r round(auc_val, 3)`. The model outperforms a majority-class baseline (`r round(baseline_accuracy * 100, 1)`% accuracy), though recall for the high-engagement class remains moderate due to class imbalance.

## Data Inspection

The dataset contains 100 observations and 7 variables. The index column (`S.No`) was removed as it carries no analytical information, leaving 6 variables for analysis.

```{r data-overview-table}
overview_df <- data.frame(
  Variable = c("USERNAME", "Followers", "Caption", "Hashtags", "Hours.since.posted", "Likes"),
  Type     = c("Categorical", "Numeric", "Text", "Text", "Numeric", "Numeric"),
  Description = c(
    "Instagram account identifier",
    "Number of followers at time of post",
    "Post caption text",
    "Hashtags used in the post",
    "Hours elapsed since post publication",
    "Number of likes received"
  )
)

kable(overview_df,
      caption = "Variable Overview — Instagram Reach Dataset",
      col.names = c("Variable", "Type", "Description")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE)
```

## Feature Engineering

Two new categorical variables were derived from the numeric data:

**`time`** — categorizes `Hours.since.posted` into three groups (2, 3, or 4+ hours), providing a discrete representation of posting age suitable for group comparisons.

**`like`** — divides `Likes` into three engagement tiers using the first and third quartiles (Q1 = `r Q1`, Q3 = `r Q3`): low (below Q1), medium (Q1–Q3), and high (above Q3).

**`high_like`** — a binary target variable: 1 if a post falls in the top like-tier, 0 otherwise.

```{r feature-tables}
time_df <- as.data.frame(table(instagram_reach$time))
colnames(time_df) <- c("Time Category", "Count")
time_df[["Relative Frequency"]] <- round(time_df$Count / sum(time_df$Count), 2)

kable(time_df,
      caption = "Distribution of Time Category Variable") %>%
  kable_styling(full_width = FALSE, position = "center")

high_like_df <- as.data.frame(table(instagram_reach$high_like))
colnames(high_like_df) <- c("Engagement Class", "Count")
high_like_df[["Relative Frequency"]] <- round(high_like_df$Count / sum(high_like_df$Count), 2)

kable(high_like_df,
      caption = "Distribution of Binary Engagement Target (high_like)") %>%
  kable_styling(full_width = FALSE, position = "center")
```

The target variable is imbalanced: approximately `r round(mean(instagram_reach$high_like == "Low/Medium Engagement") * 100)`% of posts fall in the Low/Medium Engagement class. This imbalance is an important consideration when interpreting classification performance.

## Descriptive Statistics

```{r descriptive-stats}
numeric_data <- instagram_reach[, c("Followers", "Hours.since.posted", "Likes")]

desc_table <- data.frame(
  Variable = names(numeric_data),
  Mean     = sapply(numeric_data, mean),
  SD       = sapply(numeric_data, sd),
  Min      = sapply(numeric_data, min),
  Median   = sapply(numeric_data, median),
  Max      = sapply(numeric_data, max)
)

kable(desc_table,
      digits = 2,
      format.args = list(big.mark = ","),
      caption = "Descriptive Statistics — Numeric Variables",
      col.names = c("Variable", "Mean", "Std. Dev.", "Min", "Median", "Max")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE)
```

`Followers` shows the highest variability — its standard deviation is large relative to the mean, driven by a small number of high-follower accounts. `Hours.since.posted` and `Likes` display more moderate spread.

## Distribution Analysis

The `Followers` variable is strongly right-skewed, as confirmed by both the histogram and the Q-Q plot. Most accounts have relatively few followers, but several outliers with very high follower counts pull the distribution to the right.

```{r hist-followers}
ggplot(instagram_reach, aes(x = Followers)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20,
                 fill = "#4C9BE8", color = "white", alpha = 0.85) +
  geom_density(color = "#1A3A5C", linewidth = 1) +
  labs(title = "Follower Count Distribution Across Accounts",
       subtitle = "Histogram with kernel density estimate",
       x = "Followers", y = "Density",
       caption = "Source: Instagram Reach Dataset") +
  theme_report()
```

```{r qq-followers}
ggplot(instagram_reach, aes(sample = Followers)) +
  stat_qq(color = "#4C9BE8", alpha = 0.7) +
  stat_qq_line(color = "#C0392B", linewidth = 1) +
  labs(title = "Normal Q-Q Plot — Followers",
       subtitle = "Departure from the reference line confirms non-normality",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_report()
```

## Time-Based Engagement Analysis

Posts in the highest time category (4+ hours since posting) tend to accumulate more likes, which is consistent with the general positive relationship between posting age and engagement explored in the next section.

```{r bar-time}
freq_time_df <- as.data.frame(table(instagram_reach$time))
colnames(freq_time_df) <- c("time", "count")

ggplot(freq_time_df, aes(x = time, y = count, fill = time)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  geom_text(aes(label = count), vjust = -0.5, fontface = "bold") +
  labs(title = "Share of Posts by Time Category",
       x = "Time Category (Hours Since Posting)", y = "Number of Posts") +
  theme_report()
```

```{r box-likes-by-time}
ggplot(instagram_reach, aes(x = time, y = Likes, fill = time)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Distribution of Likes Across Time Categories",
       subtitle = "Boxplot with individual observations overlaid",
       x = "Time Category", y = "Likes") +
  theme_report() +
  theme(legend.position = "none")
```

## Correlation and Bivariate Analysis

```{r correlation-matrix}
cor_matrix <- cor(numeric_data)

kable(cor_matrix,
      digits = 3,
      caption = "Pearson Correlation Matrix — Numeric Variables") %>%
  kable_styling(full_width = FALSE, position = "center")
```

There is a moderate positive correlation (r ≈ `r round(cor(instagram_reach$Hours.since.posted, instagram_reach$Likes), 2)`) between `Hours.since.posted` and `Likes`, suggesting that posts accumulate more likes as time passes. The relationship between `Followers` and `Likes` is weaker (r ≈ `r round(cor(instagram_reach$Followers, instagram_reach$Likes), 2)`), indicating that follower count is not the primary driver of engagement in this dataset.

```{r scatter-colored}
ggplot(instagram_reach, aes(x = Followers, y = Likes, color = time)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Likes vs Follower Count by Time Category",
       subtitle = paste0("Pearson r(Followers, Likes) = ",
                         round(cor(instagram_reach$Followers, instagram_reach$Likes), 3)),
       x = "Followers", y = "Likes", color = "Time Category") +
  scale_x_continuous(labels = scales::comma) +
  theme_report()
```

## Categorical Association Analysis

The contingency table and grouped bar chart below show how like categories distribute across time groups. Posts in the highest time category account for a greater share of high-like outcomes.

```{r contingency-table}
cont_table <- table(instagram_reach$time, instagram_reach$like)
cont_df <- as.data.frame.matrix(cont_table)
cont_df$time <- rownames(cont_df)
rownames(cont_df) <- NULL
cont_df <- cont_df[, c("time", "1", "2", "3")]

kable(cont_df,
      col.names = c("Time Category", "Low Likes", "Medium Likes", "High Likes"),
      caption = "Contingency Table: Time Category vs Like Category") %>%
  kable_styling(full_width = FALSE, position = "center")
```

```{r grouped-bar}
cont_df_long <- as.data.frame(cont_table)
colnames(cont_df_long) <- c("time", "like", "count")

ggplot(cont_df_long, aes(x = like, y = count, fill = time)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  geom_text(aes(label = count), position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.5) +
  labs(title = "Distribution of Like Categories Across Time Groups",
       x = "Like Category", y = "Number of Posts", fill = "Time Category") +
  theme_report()
```

## Logistic Regression Modeling

A binary logistic regression model was trained to predict whether a post achieves high engagement (`high_like = 1`), using `Followers` and `Hours.since.posted` as predictors. The dataset was split into a training set (70 posts) and a held-out test set (30 posts) to allow out-of-sample evaluation.

```{r model-coefficients}
coef_table <- as.data.frame(summary(model_full)$coefficients)
coef_table$Variable <- rownames(coef_table)
rownames(coef_table) <- NULL
coef_table <- coef_table[, c("Variable", "Estimate", "Std. Error", "z value", "Pr(>|z|)")]

kable(coef_table,
      digits = 4,
      col.names = c("Variable", "Estimate", "Std. Error", "z value", "p-value"),
      caption = "Logistic Regression Coefficients (fitted on full dataset for display)") %>%
  kable_styling(full_width = FALSE, position = "center")
```

`Hours.since.posted` shows a statistically significant positive effect (p < 0.01): as posting age increases, so does the probability of a post being classified as high-engagement. The effect of `Followers` is weaker and less consistent across observations, confirming that posting age is the most statistically significant predictor in this model.

## Model Evaluation

### Confusion Matrix

```{r confusion-matrix}
kable(cm,
      caption = "Confusion Matrix — Logistic Regression (Test Set, n = 30)") %>%
  kable_styling(full_width = FALSE, position = "center")
```

The model identifies `r tp` true high-engagement posts in the test set. It misses `r fn` high-engagement cases, which explains the relatively modest recall for that class. This pattern is expected given the class imbalance in the data.

### Performance Metrics

```{r model-metrics}
metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "F1-Score", "AUC",
             "Baseline Accuracy"),
  Value  = round(c(accuracy, precision, recall, f1, auc_val, baseline_accuracy), 4)
)

kable(metrics_df,
      caption = "Model Performance Metrics (Test Set)",
      col.names = c("Metric", "Value")) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(6, italic = TRUE, color = "grey50")
```

Although overall accuracy is `r round(accuracy * 100, 1)`%, recall for the high-engagement class remains moderate, which indicates that the model misses a non-trivial share of truly high-engagement posts. The model outperforms a majority-class baseline (`r round(baseline_accuracy * 100, 1)`% accuracy), confirming that the predictors carry genuine information — but the performance gain is limited by dataset size and class imbalance.

### ROC Curve

```{r roc-curve}
ggroc(roc_obj, color = "#E74C3C", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "grey50") +
  labs(title = "ROC Curve — Logistic Regression Model",
       subtitle = paste0("AUC = ", round(auc_val, 3),
                         "  |  Evaluated on held-out test set (n = 30)"),
       x = "Specificity", y = "Sensitivity") +
  theme_report()
```

An AUC of `r round(auc_val, 3)` indicates that the model discriminates between high and low/medium engagement posts better than chance, but there remains meaningful room for improvement, particularly in identifying the minority class.

## Conclusions

- `Hours.since.posted` is the most statistically significant predictor in this model: older posts tend to accumulate more likes.
- The relationship between `Followers` and `Likes` is positive but moderate, suggesting that follower count alone does not reliably predict engagement.
- The `Followers` distribution is strongly right-skewed, driven by a small number of high-follower accounts that inflate the mean.
- Posts in the highest time category show a higher proportion of high-like outcomes, consistent with the regression findings.
- Evaluated on a held-out test set, the logistic regression achieves `r round(accuracy * 100, 1)`% accuracy and AUC `r round(auc_val, 3)`, outperforming a naive baseline but with recall for the high-engagement class remaining the primary performance gap.

## Limitations

- **Small dataset:** With only 100 observations, model estimates are sensitive to sampling variation and results may not generalize reliably to other contexts.
- **Class imbalance:** Approximately 77% of posts fall in the low/medium engagement category, which inflates accuracy as a standalone metric and limits recall for the minority class.
- **Limited feature set:** Only two predictors are included. Additional features such as caption length, hashtag count, or posting time-of-day could improve predictive performance.
- **Single train/test split:** Performance estimates are based on one 70/30 split. Cross-validation would yield more stable evaluation at the cost of reduced interpretability at this dataset size.
